---
title: "Practical Machine Learning"
author: "Ckchan"
date: "Saturday, August 22, 2015"
output: html_document
---

##Introduction
In this project, the goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. The training data consists of accelerometer data and a label - Classe, identifying the quality of the activity the participant was doing . The testing data consists of accelerometer data without the Classe label. Thus the goal is to predict the labels for the test set observations.

##Data Preprocessing

```{r}
library(caret)
ptrain <- read.csv("pml-training.csv")
ptest <- read.csv("pml-testing.csv")
```

In order to estimate the out-of-sample error, do a random split of the full training data ptrain into a smaller training set, ptrain1 and a validation set, ptrain2

```{r}
set.seed(777)
inTrain <- createDataPartition(y=ptrain$classe, p=0.7, list=F)
ptrain1 <- ptrain[inTrain, ]
ptrain2 <- ptrain[-inTrain, ]
```

Next, reduce the number of features by removing: 
1. variables that are almost always NA 
2. variables with nearly zero variance
3. variables that do not make sense for prediction

```{r}
#remove variables that are almost always NA
mostlyNA <- sapply(ptrain1, function(x) mean(is.na(x))) > 0.95
ptrain1 <- ptrain1[, mostlyNA==F]
ptrain2 <- ptrain2[, mostlyNA==F]

#remove variables with nearly zero variance
zerovar <- nearZeroVar(ptrain1)
ptrain1 <- ptrain1[, -zerovar]
ptrain2 <- ptrain2[, -zerovar]

#remove variables that don't make much sense for prediction which are column X, user_name, raw_timestamp_part_1, raw_timestamp_part_2, cvtd_timestamp, ie, the first 5 variables
ptrain1 <- ptrain1[, -(1:5)]
ptrain2 <- ptrain2[, -(1:5)]
```

##Model Building
Start with a Random Forest model, to see if it would have acceptable performance. Fit the model on ptrain1, and instruct the "train" function to use 5-fold cross validation.

```{r}
# instruct train to use 5-fold cross validation 
fitControl <- trainControl(method="cv", number=5, verboseIter=F)

# fit model on ptrain1
fit <- train(classe ~ ., data=ptrain1, method="rf", trControl=fitControl)

# print final model to see chosen tuning parameters 
fit$finalModel

```

##Model Selection & Evaluation

The fitted model is used to predict the label, Classe in ptrain2. The confusion matrix compares the predicted vs the actual labels

```{r}
#use model to predict classe in validation set ptrain2
pred1 <- predict(fit, newdata=ptrain2)

#display confusion matrix to get estimate of out-of-sample error
confusionMatrix(ptrain2$classe, pred1)

```
The accuracy is 99.8%, so the predicted accuracy for the out-of-sample error is 0.2%.

This is an excellent result, so instead of using other algorithms, Random Forests will be used for prediction on the test set.

##Retraining the selected model
Before carrying out predictions on the test set, it is important to train the model on the full training set - ptrain, instead of a model trained on the reduced training set - ptrain1 in order to obtain the most accurate predictions. Thus, all the steps that was carry out in the preprocessing stage will be done on ptrain and ptest this time round

```{r}
#remove variables that are almost always NA
mostlyNA <- sapply(ptrain, function(x) mean(is.na(x))) > 0.95
ptrain <- ptrain[, mostlyNA==F]
ptest <- ptest[, mostlyNA==F]

#remove variables with nearly zero variance
zerovar  <- nearZeroVar(ptrain)
ptrain <- ptrain[, -zerovar ]
ptest <- ptest[, -zerovar ]

#remove variables that don't make much sense for prediction which are column X, user_name, raw_timestamp_part_1, raw_timestamp_part_2, cvtd_timestamp, ie, the first 5 variables
ptrain <- ptrain[, -(1:5)]
ptest <- ptest[, -(1:5)]

# refit model using full training set - ptrain
fitControl <- trainControl(method="cv", number=5, verboseIter=F)
fit <- train(classe ~ ., data=ptrain, method="rf", trControl=fitControl)
```

##Making predictions on the test set given

Use the model fit on ptrain to predict the label for the observations in ptest and write those predictions to individual files

```{r}
# predict on test set
preds <- predict(fit, newdata=ptest)

# convert predictions to character vector
preds <- as.character(preds)

# create function to write predictions to files
pml_write_files <- function(x) {
    n <- length(x)
    for(i in 1:n) {
        filename <- paste0("problem_id_", i, ".txt")
        write.table(x[i], file=filename, quote=F, row.names=F, col.names=F)
    }
}

# create prediction files to submit
pml_write_files(preds)
```
